Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2020-06-22 bbbq_article.tex
Improve

Transmembrane helices (TMHs) in the human proteome
are an overrepresented potential source of epitopes on major 
histocompatibility complex (MHC) class I for the majority of HLA-I haplotypes. 
It is unknown if the immune system is as able to detect 
TMHs as found in pathogens.
Additionally, it is unknown if MHC-II is also likelier to detect
human and pathogen TMHs than expected by chance.
It is unknown if natural selection is the cause that 
the immune system is more attentive for TMHs
This study shows that MHC-I [also has/does not have] more
epitopes derived from a TMH for a pathogen proteome, when compared with
a host proteome.
Additionally, MHC-II binds to polypeptides derived from TMHs 
[less/equally/more] often than expected by chance.
Lastly, using an innovative computation method, 
we show that natural selection on TMHs is [strong enough/still too weak]
to detect.
Our findings suggest that the immune system is [less/neutral/more]
vigilant to TMHs than expected by chance and this has 
left [a clear/a weak/no]
signal in the evolutionary history of the pathogen.

@@

{\bf Keywords:} antigen presentation, membrane proteins, bioinformatics, 
adaptive immunity, transmembrane domain, epitopes, T lymphocyte, 
MHC-1, MHC-I, MHC-2, MHC-II, COVID-19

@@

All nucleated cells in humans present randomly sampled polypeptides
fragments to the surroundings of the cell using the Major 
Histocompatibility Complex (MHC) class I molecules.
If a cell gets infected by a virus, also the virus' polypeptides
will be presented at the cell surface. The foreign viral antigen is detected 
by cytotoxic T lymphocytes, which will kill the infected cell.

@@

All pathogens can be detected by the foreign proteins on 
their (bacterial or fungal) cell walls or (viral) envelope.
For bacteria and fungi, this is the main mode of their detection,
as these do not infect a cell with their (foreign) DNA.
T and B cells express MHC-II to detect foreign polypeptide fragments.
An immune response is started when MHC-II detects a pathogen.

@@

Any human's immune system detects only a fraction of all possible
polypeptide fragments. 
Human MHCs are also called HLAs ('Human Leukocyte Antigens') and
because this research uses human hosts only, we will use these terms
interchangeably.
Already each MHC molecule can only bind a, possible exclusive,
subset of all possible polypeptides.
For example, HLA-A and HLA-B have no overlap in which
peptides they bind (\cite{lund2004definition}).
The HLA region of humans is highly polymorphic, 
making it hard to classify all of the many haplotypes.
Classification of HLAs is based on algorithms that
maximize the information content of each 
classification, such as the presence of a certain amino acid at 
the first (for MHC-II, \cite{southwood1998several})
or second position (for MHC-I, \cite{lund2004definition}) of a polypeptide.
However, this does not imply that all polypeptides for a classification
pattern bind to the MHC. Instead, software should be used that are
tailored to predict if a polypeptide binds to an HLA.

@@

It is helpful to be able to predict which polypeptides are immunogenic,
in, among others, vaccine development \richel{Reference here}. 
Already for a decade, synthesized polypeptide fragments are used 
to experimentally determine which polypeptides
are immunogenic \richel{Reference here}.
This approach, however, is tedious and costly.
It should come as no surprise that these results would be used
to make in silico predictions.
It may come as a surprise, however, that these predictions
are actually reliable in practice \cite{larsen2010identification,schellens2008unanticipated,tang2011genome}.

@@

For MHC-I, there are multiple computational tools developed 
to predict epitopes. 
According to \cite{lundegaard2011prediction}, in 2011,
from a set of multiple tools, 
NetMHCcons \cite{karosiene2012netmhccons} gave the best predictions.
A tool developed later is \verb;epitope-prediction; \cite{bianchi2017},
which uses a stabilized matrix method \cite{kim2009derivation}.
In this study, we will use \verb;epitope-prediction; \cite{bianchi2017}.

@@

An increased detection
is obtained by expressing a wide variety of
MHCs, which is assured by the human leukocyte antigen gene complex (HLA).
However, a different HLA will result in a different variety of MHCs,
that will display different polypeptide fragments. Additionally, there
may have been selection the HLA to display and or detect different
polypeptide fragments.

@@

It is important that MHC-I presents both the polypeptides of the 
(healthy) cell, as well as possible pathogen-derived polypeptides when the
cell is infected. As described above, MHC-I presents TMH-derived epitopes 
from the human host more often. It is unknown if MHC-I has the same
dedication to present epitopes that stem from TMHs derived from
proteins produced by pathogen, either would the pathogen
be a virus \richel{$\mathcal{H}_{1,1}$} or 
a bacterium \richel{$\mathcal{H}_{1,2}$}.

@@

TMHs, on the other hand, also have their function constraints, 
yet can occur multiple time a pathogen's proteome.
One can safely assume a pathogen's proteome contains multiple TMHs.
Therefore, it may be beneficial for the host
if its immune system would be more attentive towards TMHs.
And maybe this has already happened: MHC-I already detects hydrophobic
polypeptides. This feature, however, may also be caused by selection
to detect hydrophobic regions in the soluble proteins of pathogens.
It is unknown, when focusing on TMHs only, if a signal of selection
can be detected.

@@

To detect an evolutionary signal for selection on binding
TMHs polypeptides, we need a phylogenetic tree with 
many time-dated proteomes.
For humans, we use the 25000 proteomes used in influenza \richel{citation here}.
For viruses, we use the many time-dated COVID-19 proteomes.
We can assume that only after the moment that COVID-19 spilled over 
to humans, it has mostly been the human immune system selecting against
its detection, although this is only part of the selection for 
having a high transmission rate.
For bacteria, use the proteomes of Mycobacterium 
\richel{Really? Pick better set, if possible!}.

@@

To determine the percentages of MHC-II epitopes overlap, 
we do the same analysis as MHC-I, except for
MHC-II haplotypes. We picked the MHC-II alleles
that occur with a frequency of at least 20\% in
the human population (\cite{greenbaum2011functional},
see \ref{table:mhc2_haplotypes} for coverage),
resulting in \richel{approx 8} HLAs.
The threshold of 20\% is arbitrarily, yet chosen to result in 5 and 10 HLA variants.  
As \cite{bianchi2017}, we define the 5\% of polypeptides 
with the lowest IC50 values as binders. We then counted the percentage
of binders that have one or more amino acids that are part of a TMH.

@@

To calculate the probability all of our observations happened by chance,
we view our experiment as a two-state Markov chain, as depicted in
figure \ref{fig:markov_chain}. The two nodes depict the
two possible detection state a polypeptide is in, which is
either undetected $U$, or detected $D$ by the immune system.
The $D$ can be either detection by an MHC-I or MHC-II haplotype.
The edges depict the probability of a state transition by mutation,
for example, $p_{DU}$ denotes the probability that a detected
polypeptide mutates into an undetected one.
We focus on polypeptides that are TMH and we assume a mutation that 
renders the TMH non-functional as lethal.

@@

When we have calculated the probability of each transition (as described below),
we can simply count the number of observed transitions.
Because the probability of transitions differ per polypeptide,
the total system has the properties of a Poisson binomial distribution,
from which we can calculate the probability of the observing our
number of transitions.

@@

\paragraph{Calculating one polypeptide's probability}

We calculate the probability of a transition per polypeptide.
For any polypeptide, we can generate all sequences that differ
in only one amino acids. Of these sequences, we only keep the ones
that result in a TMH. Of each of the remaining TMH sequences, we predict
which mutants are detected by the immune system. Because not all 
AA mutations are equally likely, we correct for the AA transition
rates using the FLU transition matrix (\cite{dang2010flu}),
which is a transition matrix derived from observed influenza AA mutations.

@@

We assume that if a polypeptide binds to an MHC, it is always presented and always
leads to an immune response.
The more often this assumption is violated, 
the more our analysis will be weakened.
Experiments in mice with MHC-I epitopes, however, indicate that (if the dose 
administered was high enough) there are either none or rare
exceptions \cite{sette1994relationship}.

@@

All data sets consists out of $10^4$ observations.
Each observation starts with a randomly generated source polypeptide fragment,
as well a mutant. For the 'null' data set, the mutant is randomly select.
For the 'rigged' data set, beneficial mutations are twice as likely to
be selected. We define a beneficial mutation as a transition from
being detected to being undetected.

@@

We concluded that the
epitopes that MHC-I presents are [as/not as] likely 
to be derived from TMH within either a human host and its bacterial pathogen.
Because a bacterium does not infect a cell, thus its polypeptides
will not be presented by MHC-I, this result is [unexpected/expected]

@@

\begin{figure}[!htbp]
  \includegraphics[width=0.9\textwidth]{ic50s_per_haplotype/ic50s_per_haplotype.png}
  Predicted IC50s (in nM) per MHC-I haplotype
  \caption{
    \richel{Used randomly simulated polypeptides, results are real}
  }
  \label{fig:ic50s_per_haplotype}
\end{figure}

@@

\begin{figure}[!htbp]
  \includegraphics[width=0.9\textwidth]{p_bind_per_hydrophobicity/binds_mhc1_vs_binds_mhc2.png}
  \caption{
    \richel{Used randomly simulated polypeptides, results are real}
  }
  \label{fig:binds_mhc1_vs_binds_mhc2}
\end{figure}

\begin{figure}[!htbp]
  \includegraphics[width=0.9\textwidth]{p_bind_per_hydrophobicity/hydrophobicity_vs_binds_mhc1.png}
  \caption{
    \richel{Used randomly simulated polypeptides, results are real}
  }
  \label{fig:hydrophobicity_vs_binds_mhc1}
\end{figure}

\begin{figure}[!htbp]
  \includegraphics[width=0.9\textwidth]{p_bind_per_hydrophobicity/hydrophobicity_vs_binds_mhc2.png}
  \caption{
    \richel{Used randomly simulated polypeptides, results are real}
  }
  \label{fig:hydrophobicity_vs_binds_mhc2}
\end{figure}

\begin{figure}[!htbp]
  \includegraphics[width=0.9\textwidth]{p_bind_per_hydrophobicity/hydrophobicity_vs_is_tmh.png}
  \caption{
    \richel{Used randomly simulated polypeptides, results are real}
  }
  \label{fig:hydrophobicity_vs_is_tmh}
\end{figure}

\begin{figure}[!htbp]
  \includegraphics[width=0.9\textwidth]{p_bind_per_hydrophobicity/is_tmh_vs_binds_mhc1.png}
  \caption{
    \richel{Used randomly simulated polypeptides, results are real}
  }
  \label{fig:is_tmh_vs_binds_mhc1}
\end{figure}

\begin{figure}[!htbp]
  \includegraphics[width=0.9\textwidth]{p_bind_per_hydrophobicity/is_tmh_vs_binds_mhc2.png}
  \caption{
    \richel{Used randomly simulated polypeptides, results are real}
  }
  \label{fig:is_tmh_vs_binds_mhc2}
\end{figure}

>>>>

The human immune system is unexpectedly attentive towards
human transmembrane helices (TMHs), as TMH-derived peptide fragments
bind to histocompatibility complex (MHC) class I more often then expected
by chance only.
This finding hints that there has been selection upon detecting TMH-derived
peptides in pathogens.
If this signature of selection is present in MHC-II is unknown.
This study shows that MHC-I [also has/does not have] more
epitopes derived from a TMH for a pathogen proteome, when compared with
a host proteome.
Additionally, MHC-II binds to peptides derived from TMHs 
[less/equally/more] often than expected by chance.
Lastly, using an innovative computation method, 
we show that natural selection on TMHs is [strong enough/still too weak]
to detect.
Our findings suggest that the immune system is [less/neutral/more]
vigilant to TMHs than expected by chance and this has 
left [a clear/a weak/no]
signal in the evolutionary history of the pathogen.

@@

{\bf Keywords:} antigen presentation, membrane proteins, bioinformatics, 
adaptive immunity, transmembrane domain, transmembrane helix, 
epitopes, T lymphocyte, MHC-1, MHC-I, MHC-2, MHC-II, COVID-19

@@

All nucleated cells in humans present randomly sampled peptides
fragments to the surroundings of the cell using the Major 
Histocompatibility Complex (MHC) class I molecules.
If a cell gets infected by a virus, also the virus' peptides
will be presented at the cell surface. The foreign viral antigen is detected 
by cytotoxic T lymphocytes, which will kill the infected cell.

@@

All pathogens can be detected by the foreign proteins on 
their (bacterial or fungal) cell walls or (viral) envelope.
For bacteria and fungi, this is the main mode of their detection,
as these do not infect a cell with their (foreign) DNA.
T and B cells express MHC-II to detect foreign peptide fragments.
An immune response is started when MHC-II detects a pathogen.

@@

Any human's immune system detects only a fraction of all possible
peptide fragments.
For humans, the MHC proteins are encoded by the
HLA ('Human Leukocyte Antigens') genes.
Each MHC complex can only bind a subset of all possible peptides.
For example, HLA-A and HLA-B have no overlap in which
peptides they bind (\cite{lund2004definition}).
The HLA region of humans is highly polymorphic, 
making it hard to classify all of the many haplotypes.
Classification of HLAs is based on algorithms that
maximize the information content of each 
classification, such as the presence of a certain amino acid at 
the first (for MHC-II, \cite{southwood1998several})
or second position (for MHC-I, \cite{lund2004definition}) of a peptide.
However, this does not imply that when a peptide follows
a classification pattern, that it will bind to the MHC. 
Instead, software should be used to predict if the
full peptide will do so.

@@

It is helpful to be able to predict which peptides are immunogenic,
in, among others, vaccine development \richel{Reference here}. 
Already for a decade, synthesized peptide fragments are used 
to experimentally determine which peptides
are immunogenic \richel{Reference here}.
This approach, however, is tedious and costly.
Due to this, software was written that allows for silico 
predictions, which are reliable in practice \cite{larsen2010identification,schellens2008unanticipated,tang2011genome}.

@@

For MHC-I, there are multiple computational tools developed 
to predict epitopes. 
According to \cite{lundegaard2011prediction}, in 2011,
from a set of multiple tools, 
NetMHCcons \cite{karosiene2012netmhccons} gave the best predictions.
A tool developed later is \verb;epitope-prediction; \cite{bianchi2017},
which uses a stabilized matrix method \cite{kim2009derivation}.
The most recent FOSS tool available is MHCnuggets\cite{shao2020high},
which we will use in this study

@@

An increased detection
is obtained by expressing a wide variety of
MHCs, which is assured by the human leukocyte antigen gene complex (HLA).
However, a different HLA will result in a different variety of MHCs,
that will display different peptide fragments. Additionally, there
may have been selection the HLA to display and or detect different
peptide fragments.

@@

It is important that MHC-I presents both the peptides of the 
(healthy) cell, as well as possible pathogen-derived peptides when the
cell is infected. As described above, MHC-I presents TMH-derived epitopes 
from the human host more often. It is unknown if MHC-I has the same
dedication to present epitopes that stem from TMHs derived from
proteins produced by pathogen, either would the pathogen
be a virus \richel{$\mathcal{H}_{1,1}$} or 
a bacterium \richel{$\mathcal{H}_{1,2}$}.

@@

TMHs, on the other hand, also have their function constraints, 
yet can occur multiple time a pathogen's proteome.
One can safely assume a pathogen's proteome contains multiple TMHs.
Therefore, it may be beneficial for the host
if its immune system would be more attentive towards TMHs.
And maybe this has already happened: MHC-I already detects hydrophobic
peptides. This feature, however, may also be caused by selection
to detect hydrophobic regions in the soluble proteins of pathogens.
It is unknown, when focusing on TMHs only, if a signal of selection
can be detected.

@@

To detect an evolutionary signal for selection on binding
TMHs peptides, we need a phylogenetic tree with 
many time-dated proteomes.
For humans, we use the 25000 proteomes used in influenza \richel{citation here}.
For viruses, we use the many time-dated COVID-19 proteomes.
We can assume that only after the moment that COVID-19 spilled over 
to humans, it has mostly been the human immune system selecting against
its detection, although this is only part of the selection for 
having a high transmission rate.
For bacteria, use the proteomes of Mycobacterium 
\richel{Really? Pick better set, if possible!}.

@@

To determine the percentages of MHC-II epitopes overlap, 
we do the same analysis as MHC-I, except for
MHC-II haplotypes. We picked the MHC-II alleles
that occur with a frequency of at least 20\% in
the human population (\cite{greenbaum2011functional},
see \ref{table:mhc2_haplotypes} for coverage),
resulting in \richel{approx 8} HLAs.
The threshold of 20\% is arbitrarily, yet chosen to result in 5 and 10 HLA variants.  
As \cite{bianchi2017}, we define the 5\% of peptides 
with the lowest IC50 values as binders. We then counted the percentage
of binders that have one or more amino acids that are part of a TMH.

@@

To calculate the probability all of our observations happened by chance,
we view our experiment as a two-state Markov chain, as depicted in
figure \ref{fig:markov_chain}. The two nodes depict the
two possible detection state a peptide is in, which is
either undetected $U$, or detected $D$ by the immune system.
The $D$ can be either detection by an MHC-I or MHC-II haplotype.
The edges depict the probability of a state transition by mutation,
for example, $p_{DU}$ denotes the probability that a detected
peptide mutates into an undetected one.
We focus on peptides that are TMH and we assume a mutation that 
renders the TMH non-functional as lethal.

@@

When we have calculated the probability of each transition (as described below),
we can simply count the number of observed transitions.
Because the probability of transitions differ per peptide,
the total system has the properties of a Poisson binomial distribution,
from which we can calculate the probability of the observing our
number of transitions.

@@

\paragraph{Calculating one peptide's probability}

We calculate the probability of a transition per peptide.
For any peptide, we can generate all sequences that differ
in only one amino acids. Of these sequences, we only keep the ones
that result in a TMH. Of each of the remaining TMH sequences, we predict
which mutants are detected by the immune system. Because not all 
AA mutations are equally likely, we correct for the AA transition
rates using the FLU transition matrix (\cite{dang2010flu}),
which is a transition matrix derived from observed influenza AA mutations.

@@

We assume that if a peptide binds to an MHC, it is always presented and always
leads to an immune response.
The more often this assumption is violated, 
the more our analysis will be weakened.
Experiments in mice with MHC-I epitopes, however, indicate that (if the dose 
administered was high enough) there are either none or rare
exceptions \cite{sette1994relationship}.

@@

All data sets consists out of $10^4$ observations.
Each observation starts with a randomly generated source peptide fragment,
as well a mutant. For the 'null' data set, the mutant is randomly select.
For the 'rigged' data set, beneficial mutations are twice as likely to
be selected. We define a beneficial mutation as a transition from
being detected to being undetected.

@@

We concluded that the
epitopes that MHC-I presents are [as/not as] likely 
to be derived from TMH within either a human host and its bacterial pathogen.
Because a bacterium does not infect a cell, thus its peptides
will not be presented by MHC-I, this result is [unexpected/expected]

@@

\begin{figure}[!htbp]
  \includegraphics[width=0.9\textwidth]{ic50s_per_haplotype/ic50s_per_haplotype.png}
  Predicted IC50s (in nM) per MHC-I haplotype
  \caption{
    \richel{Used randomly simulated peptides, results are real}
  }
  \label{fig:ic50s_per_haplotype}
\end{figure}

@@

\begin{figure}[!htbp]
  \includegraphics[width=0.9\textwidth]{p_bind_per_hydrophobicity/binds_mhc1_vs_binds_mhc2.png}
  \caption{
    \richel{Used randomly simulated peptides, results are real}
  }
  \label{fig:binds_mhc1_vs_binds_mhc2}
\end{figure}

\begin{figure}[!htbp]
  \includegraphics[width=0.9\textwidth]{p_bind_per_hydrophobicity/hydrophobicity_vs_binds_mhc1.png}
  \caption{
    \richel{Used randomly simulated peptides, results are real}
  }
  \label{fig:hydrophobicity_vs_binds_mhc1}
\end{figure}

\begin{figure}[!htbp]
  \includegraphics[width=0.9\textwidth]{p_bind_per_hydrophobicity/hydrophobicity_vs_binds_mhc2.png}
  \caption{
    \richel{Used randomly simulated peptides, results are real}
  }
  \label{fig:hydrophobicity_vs_binds_mhc2}
\end{figure}

\begin{figure}[!htbp]
  \includegraphics[width=0.9\textwidth]{p_bind_per_hydrophobicity/hydrophobicity_vs_is_tmh.png}
  \caption{
    \richel{Used randomly simulated peptides, results are real}
  }
  \label{fig:hydrophobicity_vs_is_tmh}
\end{figure}

\begin{figure}[!htbp]
  \includegraphics[width=0.9\textwidth]{p_bind_per_hydrophobicity/is_tmh_vs_binds_mhc1.png}
  \caption{
    \richel{Used randomly simulated peptides, results are real}
  }
  \label{fig:is_tmh_vs_binds_mhc1}
\end{figure}

\begin{figure}[!htbp]
  \includegraphics[width=0.9\textwidth]{p_bind_per_hydrophobicity/is_tmh_vs_binds_mhc2.png}
  \caption{
    \richel{Used randomly simulated peptides, results are real}
  }
  \label{fig:is_tmh_vs_binds_mhc2}
\end{figure}
